# Introduction {#intro}





<!-- You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods). -->

<!-- Figures and tables with captions will be placed in `figure` and `table` environments, respectively. -->


<!-- ```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'} -->
<!-- par(mar = c(4, 4, .1, .1)) -->
<!-- plot(pressure, type = 'b', pch = 19) -->
<!-- ``` -->

<!-- Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab). -->

<!-- ```{r nice-tab, tidy=FALSE} -->
<!-- knitr::kable( -->
<!--   head(iris, 20), caption = 'Here is a nice table!', -->
<!--   booktabs = TRUE -->
<!-- ) -->
<!-- ``` -->

<!-- You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015]. -->



The objective of this textbook is to provide you with the shortest path to exploring your data, visualizing it, forming hypotheses and validating and defending them. Given a data set, you want to be able to make any plot you wish, find plots which show something actionable and interesting, explore data by slicing and dicing it and finally present your results in a statistically convincing manner, perhaps in a colorful and visually appealing way.

Questions which you will have to anticipate and you will have to answer are
- How do you know that your findings are not random?
- And fundamental of all questions:
- **So what?**

Even the most impressing looking results may come up randomly. And you will be asked this question along with the question *“what was your p-value and how did you compute it”*

And even if you convince your audience that your results are not random, you will have to be ready to explain why your audience should care about the results you reported. In other words, is there any actionable value in your results? Or they are just simply interesting, good to know, but no one really needs to care much about them otherwise? Hopefully it is the former not the latter.


In the following sections we will address these questions and go through the process of data exploration, validation, and presentation.

- We will start with making plots, follow with free style data exploration – which allows us to form the leads, that is hypotheses. Then we will follow with simple statistical tests which will allow us to validate these hypothesis and defend our findings against randomness claims. - We will learn how to calculate p-values and how to use them to defend our findings. 
- We will use as few R commands as possible and reach our goal in the shortest possible path. In fact we will demonstrate how using just 7 R commands we can perform quite sophisticated data exploration.  In the appendix, we show many more useful commands of R which eventually you would have to use. However, our goal in this short textbook, is to present the  shortest path to data analysis which will let you import the data, plot it, make some analysis yourself and use R-libraries.  In this textbook and in this class we do not teach how to clean the data (data wrangling) and how to deal with a wide variety of data types. We also do not address complex data transformations such as multi-frame operations like merge (we show them in appendix).  We also do not explain how different machine learning methods work, we only show you how to use them. It is similar to teaching one how to drive a car without knowing how a car engine works.

# Setting Up R 

- **Important Instructions**
  - Installation of R is required before installing RStudio
    - "R” is a programming language, and,
    - “RStudio” is an Integrated Development Environment (IDE) which provides you a platform to code in R.

- __How to download and install R & RStudio?__

  - _Downloading and installing R._

    - For Windows Users.
      - Click on the link provided below or copy paste it on your favourite browser and go to the website.
          - [https://cran.r-project.org/bin/windows/base/](https://cran.r-project.org/bin/windows/base/)
      - Click on the link at top left where it says “Download R 4.0.3 for windows” or the latest at the time of your installation.
      - Open the downloaded file and follow the instructions as it is.

    - For MAC Users.
      - Click on the link provided below or copy paste it on your favourite browser and go to the website.
          - [https://cloud.r-project.org/bin/macosx/](https://cloud.r-project.org/bin/macosx/)
      - Under “Latest release”, click on “R-4.0.3.pkg” or the latest at the time of your installation.
      - Open the downloaded file and follow the instructions as it is.
      
 
  - _Downloading and installing RStudio._
  
    - For Windows Users.
      - Click on the link below or copy paste it in your favourite browser.
          - [https://rstudio.com/products/rstudio/download/](https://rstudio.com/products/rstudio/download/)
      - Scroll down almost till the end of the web page until you find a section named “All Installers”.
      - Click on the download link beside “Windows 10/8/7” to download the windows version of RStudio.
      - Install RStudio by clicking on the downloaded file and following the instructions as it is.

    - For MAC Users.
      - Click on the link below or copy paste it in your favourite browser.
          - [https://rstudio.com/products/rstudio/download/](https://rstudio.com/products/rstudio/download/)
      - Scroll down almost till the end of the web page until you find a section named “All Installers”.
      - Click on the link beside “macOS 10.13+” to start your download the MAC version of RStudio.
      - Install RStudio by clicking on the downloaded file and following the instructions as it is.

---

# How to upload a data set?
<script src="files/js/dcl.js"></script>
```{r ,include=FALSE}
tutorial::go_interactive()

```

 - To upload the dataset/file present in csv format the read.csv() and read.csv2() functions are frequently used The read.csv() and read.csv2() have different separator symbol: for the former this is a comma, whereas the latter uses a semicolon.

- There are two options while accessing the dataset from your local machine:
  1. To avoid giving long directory paths for accessing the dataset, one should use the command **getwd()** to get the current working directory and store the dataset in the same directory. To access the dataset stored in the same directory one can use the following: **read.csv("moody2020.csv")**
  2. One can also store the dataset at a different location and can access it using the following command: (Suppose the dataset is stored inside the folder Data101_Tutorials on the desktop)
    - For Windows Users.
      - Example: **read.csv("C:/Users/Desktop/Data101_Tutorials/moody2020.csv")**

    - For MAC Users.
      - Example: **read.csv("/Users/Desktop/Data101_Tutorials/moody2020.csv")**
          
**Note: **
The directory path given here is the current working directory hosted on *Github* where the dataset has been stored.
```{r, tut=TRUE}

# Read in the data
df <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

# Print out `df`
head(df)
```

  

<!--chapter:end:chapters/intro.Rmd-->

# Plots {#plots}
<script src="files/js/dcl.js"></script>
```{r ,include=FALSE}
tutorial::go_interactive( greedy = FALSE)

```

*Lecture presentation:* <button class="btn btn-primary" data-toggle="collapse" data-target="#pred11">Plots PPT</button>
<div id="plot1" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/deeplokhande/data101demobook/main/files/pred/jeremypred1.pdf&embedded=true" width="100%" height="500px"></embed>
</div>


```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/blob/main/files/pred/Plots-2022.pdf") #web load
# head(moody)

temp<-knitr::kable(
  head(moody, 10), caption = 'Snippet of Moody Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```

<!-- --- -->

<!-- ### Topics visited in this sub-chapter -->

<!-- * Scatter Plot -->
<!-- * Barplot -->
<!-- * Boxplot -->
<!-- * Mosaic Plot -->

---

## Scatter Plot

- Scatter Plot are used to plot points on the Cartesian plane (X-Y Plane)
- Hence it is used when both the labels are numerical values.


Lets look at example of scatter plot using Moody.
```{r,tut=TRUE,height=700}
# Let's look at a 2 attribute scatter plot.
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
moody<-read.csv("https://raw.githubusercontent.com/deeplokhande/data101demobook/main/files/dataset/moody2020b.csv") #web load
plot(moody$participation,moody$score,ylab="score",xlab="participation",main=" Participation vs Score",col="red")


```
<!--

---

### Bar Plot

- A bar plot is used to plot rectangular bars proportional to the values present in a numerical vector. 
- This rectangle height is proportional to the value of the variable in the vector.
- Barplots are also used to graphically represent the distribution of a categorical variable, after converting the categorical vector into a table(i.e. frequency distribution table)
- In a bar plot, you can also give different colors to each bar.



```{r, tut=TRUE,height=700}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
moody<-read.csv("https://raw.githubusercontent.com/deeplokhande/data101demobook/main/files/dataset/moody2020b.csv") #web load
colors<- c('red','blue','cyan','yellow','green') # Assigning different colors to bars

#lets make a table for the grades of students and counts of students for each Grade. 

t<-table(moody$grade)

#once we have the table lets create a barplot for it.

barplot(t,xlab="Grade",ylab="Number of Students",col=colors,
main="Barplot for student grade distribution",border="black")
```


---

###  Box Plot

- A boxplot shows the distribution of data in a dataset. 
- A boxplot shows the following things:
  - Minimum
  - Maximum
  - Median
  - First quartile
  - Third quartile
  - Outliers
- You can create a single boxplot using just a vector or a multiple boxplot using a formula.
- When you write a formula, you should use the Tilde (~) operator. This column name on the left side of this operator goes on the y axis and the column name on the right side of this operator goes on the x axis.



```{r,tut=TRUE,height=700}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
moody<-read.csv("https://raw.githubusercontent.com/deeplokhande/data101demobook/main/files/dataset/moody2020b.csv") #web load
colors<- c('red','blue','cyan','yellow','green') # Assigning different colors to bars


#Suppose you want to find the distribution of students score per Grade. We use box plot for getting that. 
boxplot(score~grade,data=moody,xlab="Grade",ylab="Score", main="Boxplot of grade vs score",col=colors,border="black")

# the circles represent outliers.
```


<!-- ## 4. Histogram -->

<!-- Refer Slide 15. -->

<!-- ```{r} -->

<!-- #Suppose you want to find the frequecy/distribution of cars with mileage in particular range. We use histogram for this.  -->

<!-- hist(automobile$`city-mpg`,xlim = c(0,100),xlab = 'milage', main = "Histogram of Car milage",col=colors,border="black") -->

<!-- # You can Change column range using breaks. -->

<!-- ``` -->


<!-- For more detail,reference and example refer Slides -->


<!--
---
###  Mosaic Plot

- Mosaic plot is a graphical method for visualizing data from two or more qualitative variables.
- The length of the rectangles in the mosaic plot represents the frequency of that particular value.
- The width and length of the mosaic plot can be used to interpret the frequencies of the elements.
- For example, if you want to plot the number of individuals per letter grade using a smartphone, you want to look at a mosaic plot.


```{r,tut=TRUE,height=700}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
moody<-read.csv("https://raw.githubusercontent.com/deeplokhande/data101demobook/main/files/dataset/moody2020b.csv") #web load
colors<- c('red','blue','cyan','yellow','green') # Assigning different colors to bars

#suppose you want to find numbers of students with a particular grade based on their texting habits. Use Mosiac-plot.

mosaicplot(moody$grade~moody$texting,xlab = 'Grade',ylab = 'Texting habit', main = "Mosiac of grade vs texing habit in class",col=colors,border="black")


```


```
{r child="./chapters/datatransformation.Rmd"}
```
 
 -->


<!--chapter:end:chapters/plots.Rmd-->

