--- 
title: "DATA 101 Shortest Textbook"
subtitle: "How To Accomplish More With Less"
version: "Early Draft 0.1"
date: "`r Sys.Date()`"
author: 
    name: "Tomasz Imielinski"
    email: "timielinski@gmail.com"
# date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
# bibliography: [book.bib, packages.bib]
# biblio-style: apalike
# link-citations: yes
description: "This is a example of a interactive book for DATA 101 Course thought by Prof. Tomasz Imielinski (http://data101.cs.rutgers.edu/) Find the demo site at (https://dev7796.github.io/data101_tutorial/) Published using Bookdown for R."
---



# {-}

---

<!-- <script src=https://cdn.datacamp.com/datacamp-light-latest.min.js></script> -->
<script src="files/js/dcl.js"></script>


```{r ,include=FALSE}
tutorial::go_interactive( greedy = FALSE)

```

<!-- <span style="color: red;background-color: lightgrey;font-size: 25px;">NOTE: THE DATACAMP SERVER ARE CURRENTLY DOWN. PLEASE COPY PASTE THE CODE IN RSTUDIO TO RUN OFFLINE. WE'LL TRY TO GET THE SITE WORKING AGAIN SHORTLY.</span> -->
<!-- <!-- **$\color{red}{\text{{NOTE: THE DATACAMP SERVER ARE CURRENTLY DOWN. PLEASE COPY PASTE THE CODE IN RSTUDIO TO RUN OFFLINE. WE'LL TRY TO GET THE SITE WORKING AGAIN SHORTLY.}}}$** -->


This is a textbook based on the [DATA 101](http://data101.cs.rutgers.edu/) course taught by Prof. Tomasz Imielinski at Computer Science Department at Rutgers University, New Brunswick.

Data 101 is an introductory course for beginners from any field of study, interested in the field of Data Science. 

This book's pages are created using [Rmarkdown](https://rmarkdown.rstudio.com/) from Rstudio (similar to jupyter notebook) and the book is compiled using ["bookdown"](https://cran.r-project.org/web/packages/bookdown/index.html) package.

The most important aspect of this interactive book are the interactive code chunks for running code, which are powered by a minimal version of Datacamp's learning interface called "Datacamp Light". For more info visit [Datacamp Light](https://github.com/datacamp/datacamp-light)

- **Note1**: This book is undergoing constant updates following along with the course thought in Spring 2021. Topics under progress are marked with a " * ".

- **Note2**: This book uses Datacamp Light for supporting runnable code chunks. In case the code chunks do not connect to run-time session, please copy the code and run in RStudio. Also, please report if facing this issue to the instructor via [email](mailto:timielinski@gmail.com).

---





<!-- **FIRST FEW PAGES CAN BE SKIPPED. JUMP TO RECITAION 3 FOR DEMO** -->

<!-- This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$. -->

<!-- The **bookdown** package can be installed from CRAN or Github: -->

<!-- ```{r eval=FALSE} -->
<!-- install.packages("bookdown") -->
<!-- # or the development version -->
<!-- # devtools::install_github("rstudio/bookdown") -->
<!-- ``` -->

<!-- Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`. -->

<!-- To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.org/tinytex/>. -->

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
