# Code Review: Exploratory Queries in R {#cr}

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive(greedy = FALSE)

```

## Snippet 1: What is the mean imdb of low budget comedies?


```{r,tut=TRUE,height=300}
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv")

mean(movies[movies$Budget=='Low' & movies$genre=='Comedy', ]$imdb_score)

```

## Snippet 2: What is standard deviation of imdb score of  high gross Family movies?

```{r,tut=TRUE,height=300}
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv")

sd(movies[movies$Gross=='High' & movies$genre =='Family',]$imdb)

```

## Snippet 3: What is the lowest imdb score among high budget movies?

```{r,tut=TRUE,height=300}
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv")

min(movies[movies$Budget=='High',]$imdb)

```

## Snippet 4: How many low budget movies generated high gross income?

```{r,tut=TRUE,height=300}
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv")

nrow(movies[movies$Budget=='Low' & movies$Gross =='High',])

```

## Snippet 5: What is imdb score of the first non-US movie in the movies data frame?

```{r,tut=TRUE,height=300}
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv")

#You can use this simple command to quickly find out
head(movies[movies$country!='USA', ]$imdb_score)

```

## Snippet 6: What is the least frequent genre among UK movies?

```{r,tut=TRUE,height=300}
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv")

#You can use this code to find out
table(movies[movies$country=='UK',]$genre, movies[movies$country=='UK',]$country)

```

## Snippet 7: Which content rating has the lowest average imdb score?

```{r,tut=TRUE,height=300}
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv")

#You can use this code to find out
tapply(movies$imdb, movies$content, mean)

```

## Snippet 8: Movies from which country have the smallest average imdb score?

```{r,tut=TRUE,height=300}
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv")

#Better compute it, since there are too many countries for visual inspection
MA<-aggregate(movies$imdb_score, list(movies$country), mean)
colnames(MA)<-c("Country", "Mimdb")
MA<-MA[order(-MA$Mimdb), ]
MA[1,]
```

## Snippet 9: What is the least frequent genre in movies data frame?

```{r,tut=TRUE,height=300}
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv")

z<-table(movies$genre)
sort(z,decreasing=FALSE)[1]
```

## Snippet 10: z value = 2.4, whats the p-value?

```{r,tut=TRUE,height=300}

1-pnorm(2.4)

```



