# Data Frames

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive( greedy = FALSE)

```
- Data Frames are data displayed in a format as a table.
<!--
- Data Frames can have different types of data inside it. While the first column can be character, the second and third can be numeric or logical. 

- Following are the characteristics of a data frame.
  - The column names should be non-empty.
  - The row names should be unique.
  - The data stored in a data frame can be of numeric, factor or character type.
  - Each column should contain same number of data items.

Use the **data.frame()** function to create a data frame:-->

Populating the dataframe:
```{r,tut=TRUE,height=300}

# Load the dataset into the moody variable
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

# Now lets view the dataframe moody with just 5-6 tuples
head(moody)
```
<br />

- Get the summary of the dataframe

```{r,tut=TRUE,height=300}

# Load the dataset into the moody variable
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

# Use the summary() function to summarize the data from a Data Frame:
summary(moody)
```
<br />

<!--
- Just like vectors, you can access specific data **(Slicing)** in dataframes using brackets. 
- But now, instead of just using one indexing vector, we use two indexing vectors: one for the rows and one for the columns. -->
- Use the notation **data[rows, columns]**, which selects the subsets of rows and columns.

```{r,tut=TRUE,height=500}

# Load the dataset into the moody variable
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

# Return row 1
moody[1, ]

# Return column 5
moody[, 5]

# Rows 1:5 and column 2
moody[1:5, 2]

# Give me rows 1-3 and columns 2 and 4 of moody
moody[1:3, c(2:4)]
```
<!--
<br />
Use the **rbind()** and **cbind()** function to add new rows and columns in a Data Frame and use the c() function to remove rows and columns in a Data Frame:

```{r,tut=TRUE,height=500}

# Lets create 3 vectors with title, author and year.
title <- c('Data Smart','Orientalism','False Impressions','Making Software')
author <- c('Foreman, John','Said, Edward','Archer, Jeffery','Oram, Andy')
year <- c(2010,2011,2012,1998)

# Now lets create a dataframe using the above column vectors.
df <- data.frame(title, author, year)

# Add a new row
updated_df <- rbind(df, c("Computer Science", "John, Doe", 2000))

# Add a new column
updated_df <- cbind(df, ISBN = c(1.1,2.2,3.3,4.4))

# OR you can add a column as
badcol <- c(1:4)
updated_df$badcol <- badcol

# Lets look at the dataframe
updated_df

# Remove the first row and column
updated_df_final <- updated_df[-c(1), -c(1)]

# Lets look at the dataframe
updated_df_final
```
-->
<br />

- Use the **dim()** function to find the number of rows and columns in a Data Frame and you can also use the **ncol()** function to find the number of columns and **nrow()** to find the number of rows:

```{r,tut=TRUE,height=400}

# Load the dataset into the moody variable
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

# Lets look at the dimension of the dataframe
dim(moody) 

# Lets look at the length of the dataframe which is similar to ncol
length(moody)
ncol(moody)
nrow(moody)
```
<!-- --- -->

<!-- ### Topics visited in this sub-chapter -->

<!-- * Scatter Plot -->
<!-- * Barplot -->
<!-- * Boxplot -->
<!-- * Mosaic Plot -->
