# Vector

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive( greedy = FALSE)

```
- A vector is simply a list of items that are of the same type.
- To combine the list of items to a vector, use the **c()** function and separate the items by a comma.
- All of the items inside a vector to are of the same type like numerical or categorical. 


Lets look at example of creating a vector:
```{r,tut=TRUE,height=300}

#Lets create 3 vectors with title, author and year.
title <- c('Data Smart','Orientalism','False Impressions','Making Software')
author <- c('Foreman, John','Said, Edward','Archer, Jeffery','Oram, Andy')
year <- c(2010,2011,2012,1998)

#Lets look at how the created vectors look.
title
author
year

```

<br />
Create a vector with numerical values in a sequence, use the **:** operator:

```{r,tut=TRUE,height=300}

#Lets create a vectors with numerical sequence.
year <- 1:20

# Vector with numerical decimals in a sequence
weight <- 100.5 : 210.6

# To make bigger or smaller steps in a sequence, use the seq() function
height <- seq(from = 0, to = 100, by = 20)

#Lets look at how the created vectors look.
year
size
height

```

<br />
To find out how many items a vector has, use the **length()** function:

```{r,tut=TRUE,height=300}

#Lets create a vectors with categorical values
author <- c('Foreman', 'John', 'Said')

# You can access the vector items by referring to its index number inside brackets []. The first item has index 1, the second item has index 2, and so on:
author[1]

#Lets look at the size of a vectors.
length(author)

```

<br />
To sort items in a vector alphabetically or numerically, use the **sort()** function and to change the value of a specific item, refer to the **index number**:

```{r,tut=TRUE,height=300}

#Lets create a vectors with categorical values
title <- c('Data Smart','Orientalism','False Impressions','Making Software')

#sorting titles
sort(title)

# change the title 'Data Smart' to 'Smart Data'
title[1] <- "Smart Data"

#Lets look at how the updated vectors look.
title 
```

# Data Frames

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive( greedy = FALSE)

```
- Data Frames are data displayed in a format as a table.

- Data Frames can have different types of data inside it. While the first column can be character, the second and third can be numeric or logical. 

- Following are the characteristics of a data frame.
  - The column names should be non-empty.
  - The row names should be unique.
  - The data stored in a data frame can be of numeric, factor or character type.
  - Each column should contain same number of data items.

Use the data.frame() function to create a data frame:

```{r,tut=TRUE,height=700}

#Lets create 3 vectors with title, author and year.
title <- c('Data Smart','Orientalism','False Impressions','Making Software')
author <- c('Foreman, John','Said, Edward','Archer, Jeffery','Oram, Andy')

# Now lets create a dataframe using the above column vectors.
year <- c(2010,2011,2012,1998)
df <- data.frame(title, author, year)
df # Lets look at how the dataframe looks.

```

<!-- --- -->

<!-- ### Topics visited in this sub-chapter -->

<!-- * Scatter Plot -->
<!-- * Barplot -->
<!-- * Boxplot -->
<!-- * Mosaic Plot -->

---


---

# Table

- A bar plot is used to plot rectangular bars proportional to the values present in a numerical vector. 
- This rectangle height is proportional to the value of the variable in the vector.
- Barplots are also used to graphically represent the distribution of a categorical variable, after converting the categorical vector into a table(i.e. frequency distribution table)
- In a bar plot, you can also give different colors to each bar.



```{r, tut=TRUE,height=700}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
moody<-read.csv("https://raw.githubusercontent.com/deeplokhande/data101demobook/main/files/dataset/moody2020b.csv") #web load
colors<- c('red','blue','cyan','yellow','green') # Assigning different colors to bars

#lets make a table for the grades of students and counts of students for each Grade. 

t<-table(moody$grade)

#once we have the table lets create a barplot for it.

barplot(t,xlab="Grade",ylab="Number of Students",col=colors,
main="Barplot for student grade distribution",border="black")
```

