# Vector

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive( greedy = FALSE)

```
- A vector is simply a list of items that are of the same type.
- To combine the list of items to a vector, use the **c()** function and separate the items by a comma.
- All of the items inside a vector to are of the same type like numerical or categorical. 


Lets look at example of creating a vector:
```{r,tut=TRUE,height=300}

#Lets create 3 vectors with title, author and year.
title <- c('Data Smart','Orientalism','False Impressions','Making Software')
author <- c('Foreman, John','Said, Edward','Archer, Jeffery','Oram, Andy')
year <- c(2010,2011,2012,1998)

#Lets look at how the created vectors look.
title
author
year

```

<br />
Create a vector with numerical values in a sequence, use the **:** operator:

```{r,tut=TRUE,height=300}

#Lets create a vectors with numerical sequence.
year <- 1:20

# Vector with numerical decimals in a sequence
weight <- 100.5 : 210.6

# To make bigger or smaller steps in a sequence, use the seq() function
height <- seq(from = 0, to = 100, by = 20)

#Lets look at how the created vectors look.
year
size
height

```

<br />
To find out how many items a vector has, use the **length()** function:

```{r,tut=TRUE,height=300}

#Lets create a vectors with categorical values
author <- c('Foreman', 'John', 'Said')

# You can access the vector items by referring to its index number inside brackets []. The first item has index 1, the second item has index 2, and so on:
author[1]

#Lets look at the size of a vectors.
length(author)

```

<br />
To sort items in a vector alphabetically or numerically, use the **sort()** function and to change the value of a specific item, refer to the **index number**:

```{r,tut=TRUE,height=300}

#Lets create a vectors with categorical values
title <- c('Data Smart','Orientalism','False Impressions','Making Software')

#sorting titles
sort(title)

# change the title 'Data Smart' to 'Smart Data'
title[1] <- "Smart Data"

#Lets look at how the updated vectors look.
title 
```

# Data Frames

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive( greedy = FALSE)

```
- Data Frames are data displayed in a format as a table.

- Data Frames can have different types of data inside it. While the first column can be character, the second and third can be numeric or logical. 

- Following are the characteristics of a data frame.
  - The column names should be non-empty.
  - The row names should be unique.
  - The data stored in a data frame can be of numeric, factor or character type.
  - Each column should contain same number of data items.

Use the **data.frame()** function to create a data frame:

```{r,tut=TRUE,height=300}

# Lets create 3 vectors with title, author and year.
title <- c('Data Smart','Orientalism','False Impressions','Making Software')
author <- c('Foreman, John','Said, Edward','Archer, Jeffery','Oram, Andy')
year <- c(2010,2011,2012,1998)

# Now lets create a dataframe using the above column vectors.
df <- data.frame(title, author, year)
df # Lets look at how the dataframe looks.

# Use the summary() function to summarize the data from a Data Frame:
summary(df)
```

<br />
Use the **rbind()** and **cbind()** function to add new rows and columns in a Data Frame and use the c() function to remove rows and columns in a Data Frame:

```{r,tut=TRUE,height=500}

# Lets create 3 vectors with title, author and year.
title <- c('Data Smart','Orientalism','False Impressions','Making Software')
author <- c('Foreman, John','Said, Edward','Archer, Jeffery','Oram, Andy')
year <- c(2010,2011,2012,1998)

# Now lets create a dataframe using the above column vectors.
df <- data.frame(title, author, year)

# Add a new row
updated_df <- rbind(df, c("Computer Science", "John, Doe", 2000))

# Add a new column
updated_df <- cbind(df, ISBN = c(1.1,2.2,3.3,4.4))

# OR you can add a column as
badcol <- c(1:4)
updated_df$badcol <- badcol

# Lets look at the dataframe
updated_df

# Remove the first row and column
updated_df_final <- updated_df[-c(1), -c(1)]

# Lets look at the dataframe
updated_df_final
```

<br />
Use the **dim()** function to find the amount of rows and columns in a Data Frame and you can also use the **ncol()** function to find the number of columns and **nrow()** to find the number of rows:

```{r,tut=TRUE,height=400}

# Lets create 3 vectors with title, author and year.
title <- c('Data Smart','Orientalism','False Impressions','Making Software')
author <- c('Foreman, John','Said, Edward','Archer, Jeffery','Oram, Andy')
year <- c(2010,2011,2012,1998)

# Now lets create a dataframe using the above column vectors.
df <- data.frame(title, author, year)

# Lets look at the dimension of the dataframe
dim(df)

# Lets look at the length of the dataframe which is similar to ncol
length(df)
ncol(df)
nrow(df)
```
<!-- --- -->

<!-- ### Topics visited in this sub-chapter -->

<!-- * Scatter Plot -->
<!-- * Barplot -->
<!-- * Boxplot -->
<!-- * Mosaic Plot -->


# Table

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive( greedy = FALSE)

```

- Tables (i.e. frequency distribution table) can be created using **table()** along with some of its variations. 
- To use **table()**, simply add in the variables you want to tabulate separated by a comma. 
- You must reference the variable using **dataset$variable**. 
- By default, missing values are excluded from the counts; if you want a count for these missing values you must specify the argument **useNA=“ifany”** or **useNA=“always”**. 

The below examples show how to use this function.
```{r, tut=TRUE,height=500}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
moody<-read.csv("https://raw.githubusercontent.com/deeplokhande/data101demobook/main/files/dataset/moody2020b.csv") #web load

#lets make a table for the grades of students and counts of students for each Grade. 
grade_freq_table <- table(moody$grade)
texting_freq_table <- table(moody$texting)
participation_freq_table <- table(moody$participation, useNA = "ifany")

#lets see the above frequency distrbuted tables
grade_freq_table
texting_freq_table
participation_freq_table
```

