# 🔖 Free Style: Data Exploration 

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive()
```

- **Lecture slides: **     <button class="btn btn-primary" data-toggle="collapse" data-target="#freestyle12">Data Exploration</button>
<div id="freestyle12" class="collapse">    
<embed src="https://docs.google.com/presentation/d/1wvWxu_DEEoYNq0qKEkepbqu2-x0SGsXGQiE2iWbv2Iw/edit?usp=sharing" width="100%" height="500px"></embed>
</div>


## Moody Data Puzzle - Pattern I Injected

<button class="btn btn-primary" data-toggle="collapse" data-target="#MDP12">Moody Data Puzzle</button>
<div id="MDP12" class="collapse">    
<embed src="https://docs.google.com/presentation/d/1TFNngoXesu1dY5YZpGgUPOQwLWSS7IQmdZ5fejDgSxQ/edit?usp=sharing" width="100%" height="500px"></embed>
</div>

### Additional reference: Correlation and Causation

https://www.stewartmath.com/precalc_7e_dp/precalc_7e_dp6.html

## Best Student's Submissions 2022

### Presentation I
<button class="btn btn-primary" data-toggle="collapse" data-target="#PPT2021">Lauretta Martin</button>
<div id="PPT2021" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/LM-HW3.pdf&embedded=true" width="100%" height="500px"></embed>
</div>

#### Snippet 1

```{r,tut=TRUE,height=300}
moody2022 <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")


```

#### Snippet 2

```{r,tut=TRUE,height=300}
moody2022 <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")

```

#### Snippet 3

```{r,tut=TRUE,height=300}
moody2022 <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")

```

#### Snippet 4

```{r,tut=TRUE,height=300}
moody2022 <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")

```

#### Snippet 5

```{r,tut=TRUE,height=300}
moody2022 <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")

```

<br>

### Presentation II
<button class="btn btn-primary" data-toggle="collapse" data-target="#PPT2020">Sanjaya Budhathoki</button>
<div id="PPT2020" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/SB-HW3.pdf&embedded=true" width="100%" height="500px"></embed>
</div>

#### Snippet 1

Lets look at the general distribution of Grades and scores

```{r,tut=TRUE,height=400}
moody2022 <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")

#Barplot General Grade distribution
colors<- c('red','blue','cyan','yellow','green', 'orange')
barplot(table(moody2022$GRADE), col = colors, main = 'General Grade Distribution',
        ylab = 'Number of students', xlab = 'Letter Grade')

#Boxplot grade distribution by score
colors<- c('red','blue','cyan','yellow','green', 'orange')
boxplot(SCORE~GRADE, data=moody2022, xlab = "Letter Grade", ylab="Score", 
        main="Distribution of Scores by Grade", col = colors, border = "black")

```

#### Snippet 2

Does Texting affect grades?

```{r,tut=TRUE,height=600}
moody2022 <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")

#barplot using varibale always TEXTING_IN_CLASS
alwaysTexting <- moody2022[which(moody2022$TEXTING_IN_CLASS == 'always'), ]
colors<- c('red','blue','cyan','yellow','green', 'orange')
barplot(table(alwaysTexting$GRADE), col = colors, 
        main = 'Grade Distribution for student always texting', xlab = 'letter Grade',
        ylab = 'Number of Students')

#barplot using varibale sometimes TEXTING_IN_CLASS
sometimesTexting <- moody2022[which(moody2022$TEXTING_IN_CLASS == 'sometimes'), ]
colors<- c('red','blue','cyan','yellow','green', 'orange')
barplot(table(sometimesTexting$GRADE), col = colors, 
        main = 'Grade Distribution for student sometimes texting', xlab = 'letter Grade',
        ylab = 'Number of Students')

#barplot using varibale rarely TEXTING_IN_CLASS
rarelyTexting <- moody2022[which(moody2022$TEXTING_IN_CLASS == 'rarely'), ]
colors<- c('red','blue','cyan','yellow','green', 'orange')
barplot(table(rarelyTexting$GRADE), col = colors, 
        main = 'Grade Distribution for student rarely texting', xlab = 'letter Grade',
        ylab = 'Number of Students')

#barplot using varibale never TEXTING_IN_CLASS
neverTexting <- moody2022[which(moody2022$TEXTING_IN_CLASS == 'never'), ]
colors<- c('red','blue','cyan','yellow','green', 'orange')
barplot(table(neverTexting$GRADE), col = colors, 
        main = 'Grade Distribution for student never texting', xlab = 'letter Grade',
        ylab = 'Number of Students')

```

#### Snippet 3

Does Dozing off affect grades?

```{r,tut=TRUE,height=600}
moody2022 <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")

#barplot using varibale always Dozing_off
alwaysDozes <- moody2022[which(moody2022$DOZES_OFF == 'always'), ]
colors<- c('red','blue','cyan','yellow','green', 'orange')
barplot(table(alwaysDozes$GRADE), col = colors, 
        main = 'Grade Distribution for student always Dozing', xlab = 'letter Grade',
        ylab = 'Number of Students')

#barplot using varibale rarely Dozing_off
rarelyDozes <- moody2022[which(moody2022$DOZES_OFF == 'rarely'), ]
colors<- c('red','blue','cyan','yellow','green', 'orange')
barplot(table(rarelyDozes$GRADE), col = colors, 
        main = 'Grade Distribution for student rarely Dozing', xlab = 'letter Grade',
        ylab = 'Number of Students')

#barplot using varibale sometimes Dozing_off
sometimesDozes <- moody2022[which(moody2022$DOZES_OFF == 'sometimes'), ]
colors<- c('red','blue','cyan','yellow','green', 'orange')
barplot(table(sometimesDozes$GRADE), col = colors, 
        main = 'Grade Distribution for student sometimes Dozing', xlab = 'letter Grade',
        ylab = 'Number of Students')

#barplot using varibale nevers Dozing_off
neverDozes <- moody2022[which(moody2022$DOZES_OFF == 'never'), ]
colors<- c('red','blue','cyan','yellow','green', 'orange')
barplot(table(neverDozes$GRADE), col = colors, 
        main = 'Grade Distribution for student never Dozing', xlab = 'letter Grade',
        ylab = 'Number of Students')


```

#### Snippet 4

Does Particiaption affect grades?

```{r,tut=TRUE,height=500}
moody2022 <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")

#barplot using varibale 0-0.5 Participation
lessParticipation <- moody2022[which(moody2022$PARTICIPATION >= 0 & moody2022$PARTICIPATION < 0.5), ]
colors<- c('red','blue','cyan','yellow','green', 'orange')
barplot(table(lessParticipation$GRADE), col = colors, 
        main = 'Grade Distribution for student with low Participation', xlab = 'Letter Grade',
        ylab = 'Number of Students')

#barplot using varibale 0.5-1 Participation
highParticipation <- moody2022[which(moody2022$PARTICIPATION >= 0.5 & moody2022$PARTICIPATION <= 1), ]
colors<- c('red','blue','cyan','yellow','green', 'orange')
barplot(table(highParticipation$GRADE), col = colors, 
        main = 'Grade Distribution for student with high Participation', xlab = 'Letter Grade',
        ylab = 'Number of Students')
```

<br>

### Presentation III
<button class="btn btn-primary" data-toggle="collapse" data-target="#PPT2019">Sandhya Senthilkumar</button>
<div id="PPT2019" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/HW3-SS.pptx&embedded=true" width="100%" height="500px"></embed>
</div>

#### Snippet 1

Does participation affect grade or score?

```{r,tut=TRUE,height=400}
moody <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")

barplot(table(moody$GRADE), col=rainbow(5), main="Bar Plot of Grades", xlab = "Grades", ylab = "Number of Students", 
        border=NA, ylim=c(0, 200))


boxplot(moody$PARTICIPATION~moody$GRADE, main = "Participation vs. Grade", xlab = "Participation", ylab = "Grade", 
        col = rainbow(5))


boxplot(moody$SCORE~moody$GRADE, main = "Scores vs. Grades", col = rainbow(5), xlab = "Grades", 
        ylab = "Number of Students", ylim=c(0, 120))


plot(moody$PARTICIPATION, moody$SCORE, main = "Participation vs. Score", xlab = "Participation", ylab = "Score", 
     col = "blue")

```

#### Snippet 2

Does Dozing off affect grades?

```{r,tut=TRUE,height=400}
moody2022 <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")

barplot(table(moody$GRADE[moody$DOZES_OFF == "never"]), col=rainbow(5), main="Grades vs. Never Sleeping in Class", 
        xlab = "Grades", ylab = "Number of Students", border=NA, ylim=c(0, 200))

barplot(table(moody$GRADE[moody$DOZES_OFF == "sometimes"]), col=rainbow(5), main="Grades vs. Sometimes Sleeping in Class", 
        xlab = "Grades", ylab = "Number of Students", border=NA, ylim=c(0, 200))

barplot(table(moody$GRADE[moody$DOZES_OFF == "always"]), col=rainbow(5), main="Grades vs. Always Sleeping in Class", 
        xlab = "Grades", ylab = "Number of Students", border=NA, ylim=c(0, 200))

```

#### Snippet 3

Does Texting in class affect grades?

```{r,tut=TRUE,height=400}
moody2022 <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")

barplot(table(moody$GRADE[moody$TEXTING_IN_CLASS == "never"]), col=rainbow(5), main="Grades vs. Never Texting in Class", 
        xlab = "Grades", ylab = "Number of Students", border=NA, ylim=c(0, 200))

barplot(table(moody$GRADE[moody$TEXTING_IN_CLASS == "rarely"]), col=rainbow(5), main="Grades vs. Rarely Texting in Class", 
        xlab = "Grades", ylab = "Number of Students", border=NA, ylim=c(0, 200))

barplot(table(moody$GRADE[moody$TEXTING_IN_CLASS == "always"]), col=rainbow(5), main="Grades vs. Always Texting in Class", 
        xlab = "Grades", ylab = "Number of Students", border=NA, ylim=c(0, 200))

```
