# 🔖 Data puzzles

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive()
```

Data Puzzles are synthetically generated datasets with some embedded patterns.  Patterns have various forms from relationships between attributes to rules of the form “if condition then value” between specific attribute-value pairs.  These patterns are stochastic and embedded in datasets using **DataMaker** -  our  Data Puzzle Generation Tool.

We use data puzzles extensively in the class assignments.  These range from data exploration and plotting through  hypothesis testing to prediction and machine learning.  After the assignment is completed we reveal the data secrets - the patterns which were embedded by DataMaker.  Students do not have to find exactly the embedded patterns, often they find related patterns which makes the “game” even more fun. 

 In the following  we provide the list of data puzzles along with the underlying data sets.  Using DataMaker we change the patterns and even data sets from academic year to academic year..  We can also provide data puzzles of different levels of difficulty from the  one star (easy) to five star (most difficult)  ones. 



## Strange grading methods of Professor Moody Data Puzzle

Download: <a href="https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv" download="moody2022_new.csv">moody2022_new.csv</a>

How to get a good grade in Professor Moody’s class?  

Professor Moody does not give final grades just on the basis of your total score alone. Our data shows that two students with the same total score may get widely varying final grades. Can you believe that you can even fail his class with a score as high as 82%? This is outrageous, isn’t it? 

DataMaker has generated  thousands of tuples  which in addition to the total score and final grade also store bizarre information about student behaviors in the class - do they often doze off? Does a student text a lot?  Does s/he ask a lot of questions?    Does it help if you ask a lot of questions? Does it hurt if you doze off a lot?

**Comment:** There is a series of Professor Moody’s puzzles which we have used over the years. We have used different attributes including student’s major, , seniority, class participation etc. 



```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv") #web load
# head(moody)
temp<-knitr::kable(
  head(moody, 5), caption = 'Snippet of Moody Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  head(moody, 5), caption = 'Snippet of Moody Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```
<br>

```{r,tut=TRUE,height=600}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv")

summary(moody)
```

<!--
### Secrets Revealed- Patterns in Professor Moody's data? 

<button class="btn btn-primary" data-toggle="collapse" data-target="#MDP12">My Patterns</button>
<div id="MDP12" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/HW3-Moody-My-patterns.pptx&embedded=true" width="100%" height="500px"></embed>
</div>
<br>
Many student solutions falsely attribute higher grades to higher values of participation attribute.. 

This is a classic example of a hidden variable described in the reference attached below. The truth is that participation attribute value impacts the score attribute value. Generally, the higher the participation in Moody’s class, the higher the score. But it is the score attribute which has a direct impact on the grade. Thus, it is the score which is the real “hidden variable” impacting the final grade.

Thus, the score already reflects participation. Professor Moody seemed to look only at texting and dozing off attributes in grade determination (see the power points above with the explanation) 

Compare with examples of hidden variables in the following reference about correlation and causation. 

https://www.stewartmath.com/precalc_7e_dp/precalc_7e_dp6.html 

-->
<!--
### Best Student's Submissions 2022

<button class="btn btn-primary" data-toggle="collapse" data-target="#PPT2021">Lauretta Martin</button>
<button class="btn btn-primary" data-toggle="collapse" data-target="#PPT2020">Sanjaya Budhathoki</button>
<button class="btn btn-primary" data-toggle="collapse" data-target="#PPT2019">Sandhya Senthilkumar</button>

<div id="PPT2021" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/LM-HW3.pdf&embedded=true" width="100%" height="500px"></embed>
</div>

<div id="PPT2020" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/SB-HW3.pdf&embedded=true" width="100%" height="500px"></embed>
</div>

<div id="PPT2019" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/HW3-SS.pptx&embedded=true" width="100%" height="500px"></embed>
</div>
-->


## Magic Coin: Heads or Tails? Data Puzzle

Download: <a href="https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Tosses2022.csv" download="Tosses2022.csv">Tosses2022.csv</a>

Your goal is to discover secrets of a magic coin. Ordinary coins when tossed a large number of times  result equally often in Heads and in Tails.  The magic coin  though is not an ordinary coin.  Your job is to discover when the magic coin is most biased one way or another.  The data consists of thousands of tosses of the magic coin including where the toss took place, who tossed it, who reported the result as well as the time of the day.

```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
toss<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Tosses2022.csv") #web load
# head(moody)
temp<-knitr::kable(
  toss[sample(1:nrow(toss),5), ], caption = 'Snippet of Magic Coin Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  toss[sample(1:nrow(toss),5), ], caption = 'Snippet of Magic Coin Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```

<br>
```{r,tut=TRUE,height=600}
toss<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Tosses2022.csv")

summary(toss)
```

## How to predict a good party?  Data puzzle

Download: <a href="https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Partyb.csv" download="Partyb.csv">Partyb.csv</a>

```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
party<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Partyb.csv") #web load
# head(moody)
temp<-knitr::kable(
  party[sample(1:nrow(party),5), ], caption = 'Snippet of Party Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  party[sample(1:nrow(party),5), ], caption = 'Snippet of Party Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```

DataMaker has generated data about thousands of parties, some fun parties, others which were OK or simply boring.  Your goal is to discover secrets of a fun party. Is it music? Dancing? Does the host matter? Or who was present at a party? Maybe who was NOT present at the party?  All this data is stored in this data puzzle.



<br>
```{r,tut=TRUE,height=600}
party<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Partyb.csv")

summary(party)
```

## When election is truly local - data puzzle

Download: <a href="https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Voting1.csv" download="Voting1.csv">Voting1.csv</a>

In local elections in some small towns, candidates of three local parties: Royalists, KnowNothings and Anarchists are running for the office of the mayor. DataMaker has generated a survey of thousands of town residents and their political sympathies. Data of course can not be more local, leaving global concerns such as inflation or global warming to national or state office candidates.  

Here, the electorate cares  about  issues such as “should we allow leaflowers” (all, only electric, none?), what about CBD stores in town (none, just one, no restrictions), How about liquor (should the town be dry? Or hard liqueurs only). Speed limits? (none, 10mph etc) or even more extreme - the whole town being car-free, streets open only to bicycles and pedestrians?

Can we develop the profiles of voters for each of the parties? What does the  anarchist electorate  care about? Which party is leading among young people who do not want any speed limits in town?


```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
voting<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Voting1.csv") #web load
# head(moody)
temp<-knitr::kable(
  voting[sample(1:nrow(voting),5), ], caption = 'Snippet of Voting Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  voting[sample(1:nrow(voting),5), ], caption = 'Snippet of Voting Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```

<br>
```{r,tut=TRUE,height=600}
vote<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Voting1.csv")

summary(vote)
```

## Secrets of good sleep Data Puzzle 

Download: <a href="https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/SleepPrediction2.csv" download="SleepPrediction2.csv">SleepPrediction2.csv</a>

Who wouldn’t want to know the secrets of good sleep?  DataMaker has created a data set which may help to find these secrets. We store the number of exercise calories burnt during the day, the amount of wimpy tea a person has drunk (in ounces), hours spent on the computer and the quality of the preceding night’s sleep. 

```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/SleepPrediction2.csv") #web load
# head(moody)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Sleep Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Sleep Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```

<br>
```{r,tut=TRUE,height=600}
sleep<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/SleepPrediction2.csv")

summary(sleep)
```

## Let’s go to the movies:  Data Puzzle 

Download: <a href="https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv" download="Movies2022F-4.csv">Movies2022F-4.csv</a>

Using DataMaker  we have started with the imdb data set from Kaggle and embedded some patterns in it.  The original data set contains data about 12,800+ movies. We have expanded this data set by DataMaker’s opinions. Yes, only DataMaker can have an opinion on each of 12,800 movies! Can you predict which  movies does DataMaker love and which movies bore him so much that she quit?   What movies DataMaker passionately hates (hmm is DataMaker even passionate about anything at all?). 
``When does DataMaker agree with the imdb score?``

Can one predict an imdb score on the basis of a combination of DataMaker opinion (sort of super critic) and other attributes?


```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv") #web load
# head(moody)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Movies Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Movies Dataset',
  booktabs = TRUE
)
<<<<<<< HEAD
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Movies Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Movies Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Movies Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Movies Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Movies Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Movies Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Movies Dataset',
  booktabs = TRUE
)
=======
>>>>>>> a4896dd013af2182df378aac8f559becda95fd26
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```
<br>

```{r,tut=TRUE,height=600}
movie<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv")

summary(movie)
```

<!--
### Secrets Revealed- Patterns in Movies data? 

<button class="btn btn-primary" data-toggle="collapse" data-target="#MV12">Secrets Revealed</button>
<div id="MV12" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/profhw4.pdf&embedded=true" width="100%" height="500px"></embed>
</div>



### Best Student's Submissions 2022

<button class="btn btn-primary" data-toggle="collapse" data-target="#hw41">Joshua Sze</button>
<button class="btn btn-primary" data-toggle="collapse" data-target="#hw42">Andrew Fasano</button>

<div id="hw41" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/student1hw4.pptx&embedded=true" width="100%" height="500px"></embed>
</div>

<div id="hw42" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/studen2hw4.pptx&embedded=true" width="100%" height="500px"></embed>
</div>
-->

## When canvas goes wild data puzzle 

Download: <a href="https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Grades2022.csv" download="Grades2022.csv">Grades2022.csv</a>

You are all familiar with Canvas, right?  This is where you look to see your grades for each assignment and exam.  This is where you see the scores.  However it seems that Canvas went a bit wild and unfair in this data set.

``One can still fail the class with the score of 82 (sounds familiar, yes, Professor Moody would do it, but Canvas?``   How can one get a lower grade  with a higher score?  

Yes, Canvas was instructed by someone and your goal is to discover the grading method.  ``How to get an A, how to pass?``  We know who that someone is… it is **DataMaker** of course. 

```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Grades2022.csv") #web load
# head(moody)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Canvas Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Canvas Dataset',
  booktabs = TRUE
)
<<<<<<< HEAD

=======
>>>>>>> a4896dd013af2182df378aac8f559becda95fd26
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```
<br>

```{r,tut=TRUE,height=600}
canvas<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Grades2022.csv")

summary(canvas)
```



## Very local minimarket data puzzle 

Download: <a href="https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/HomeworkMarket2022.csv" download="Minimarket.csv">HomeworkMarket2022.csv</a>

What items sell together? 

A small local minimarket chain (think Wawa at its early days) has a few locations in New Jersey and it sells beer, snacks, sweets. DataMaker provided the data set of several thousand of transactions in the minimarket storing what items were purchased, when they were purchased (weekday or weekend) at which location.  

```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/HomeworkMarket2022.csv") #web load
# head(moody)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Minimarket Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Minimarket Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```
<br>

```{r,tut=TRUE,height=600}
minimarket<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/HomeworkMarket2022.csv")

summary(minimarket)
```

<!--
### What were the secret associations  between  items in the minimarket?

<button class="btn btn-primary" data-toggle="collapse" data-target="#MV12">Secrets Revealed</button>
<div id="MV12" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/ecrets-revealed-prediction-minimarket.pptx&embedded=true" width="100%" height="500px"></embed>
</div>

## Virus Data Puzzle

Whose computer  is in most danger of being infected with a virus?  Let's assume that obvious protections have been secured: antivirus software is installed and an uncrackable password is in place. 

Anything else?  But how about your personal characteristics - are you an optimist or a pessimist?  Do you have lots of facebook friends?  Is your instagram followed?  Massively?  Do you spend all your time online? Are you an outdoorsy type ? Do you have a dog? 

## Good mood Data Puzzle

What are the secrets of DataMaker’s good mood?  Now we fully admit that DataMaker is moody!  Data is not surprisingly very nerdy - amount of free disk space, number of processes, CPU utilization, number of active connections.

## Airbnb data puzzle

## Predicting grades in Professor Moody's class

### How did I cook the Professor Moody Prediction challenge data? 

<button class="btn btn-primary" data-toggle="collapse" data-target="#MV13">Secrets Revealed</button>
<div id="MV13" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/Secrets-revealed-prediction-challenge.pptx&embedded=true" width="100%" height="500px"></embed>
</div>
-->

## Addiotional Reference 

<button class="btn btn-primary" data-toggle="collapse" data-target="#p12"> Prediction - Free Style </button> 
<div id="p12" class="collapse">
<embed src="https://docs.google.com/presentation/d/1pA0bzMGr_Tu2CXsgtT9Ks5apHqxWh8l1XXBVvwWr6zc/edit?usp=sharing" width="100%" height="500px"></embed>
</div>
