# 🔖 Data puzzles

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive()
```

Data Puzzles are synthetically generated datasets with some embedded patterns.  Patterns have various forms from relationships between attributes to rules of the form “if condition then value” between specific attribute-value pairs.  These patterns are stochastic and embedded in datasets using **DataMaker** -  our  Data Puzzle Generation Tool developed by my team at Rutgers. 

We use data puzzles extensively in the class assignments.  These range from data exploration and plotting through  hypothesis testing to prediction and machine learning.  After the assignment is completed we reveal the data secrets - the patterns which were embedded by DataMaker.  Students do not have to find exactly the embedded patterns, often they find related patterns which makes the “game” even more fun. 

 In the following  we provide the list of data puzzles along with the underlying data sets.  Using DataMaker we change the patterns and even data sets from class to class.  We can also provide data puzzles of different levels of difficulty from the  one star (easy) to five star (most difficult)  ones. 


## Moody Data Puzzle

How to get a good grade in Professor Moody’s class?  

Clearly, Professor Moody does not give final grades on the basis of your total score alone. Our data shows that two students with the same total score may get widely varying final grades. Can you believe that you can even fail his class with a score as high as 82%? This is outrageous, isn’t it? 

We have collected thousands of cases over the years which in addition to the total score and final grade also store bizarre and a bit intrusive information about student behaviors in the class - do they often doze off? Do they text a lot? Ask a lot of questions?  We do not know how on earth does Moody collect this data?   Does it help if you ask a lot of questions? Does it hurt if you doze off a lot?

**Comment:** There is a series of Professor Moody’s puzzles which we have used over the years. We have used different attributes including the major of a student, seniority, participation etc. 


```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022_new.csv") #web load
# head(moody)
temp<-knitr::kable(
  head(moody, 5), caption = 'Snippet of Moody Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```
<br>

### Secrets Revealed- Patterns in Professor Moody's data? 

<button class="btn btn-primary" data-toggle="collapse" data-target="#MDP12">My Patterns</button>
<div id="MDP12" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/HW3-Moody-My-patterns.pptx&embedded=true" width="100%" height="500px"></embed>
</div>
<br>
Many student solutions falsely attribute higher grades to higher values of participation attribute.. 

This is a classic example of a hidden variable described in the reference attached below. The truth is that participation attribute value impacts the score attribute value. Generally, the higher the participation in Moody’s class, the higher the score. But it is the score attribute which has a direct impact on the grade. Thus, it is the score which is the real “hidden variable” impacting the final grade.

Thus, the score already reflects participation. Professor Moody seemed to look only at texting and dozing off attributes in grade determination (see the power points above with the explanation) 

Compare with examples of hidden variables in the following reference about correlation and causation. 

https://www.stewartmath.com/precalc_7e_dp/precalc_7e_dp6.html 


<!--
### Best Student's Submissions 2022

<button class="btn btn-primary" data-toggle="collapse" data-target="#PPT2021">Lauretta Martin</button>
<button class="btn btn-primary" data-toggle="collapse" data-target="#PPT2020">Sanjaya Budhathoki</button>
<button class="btn btn-primary" data-toggle="collapse" data-target="#PPT2019">Sandhya Senthilkumar</button>

<div id="PPT2021" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/LM-HW3.pdf&embedded=true" width="100%" height="500px"></embed>
</div>

<div id="PPT2020" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/SB-HW3.pdf&embedded=true" width="100%" height="500px"></embed>
</div>

<div id="PPT2019" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/HW3-SS.pptx&embedded=true" width="100%" height="500px"></embed>
</div>
-->


## Magic Coin Data Puzzle

Download: <a href="https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Tosses2022.csv" download="Tosses2022.csv">Tosses2022.csv</a>

Your goal is to discover secrets of a magic coin. Ordinary coins when tossed a large number of times  result equally often in Heads and in Tails.  The magic though is not an ordinary coin.  Your job is to discover when the magic coin is most biased one way or another.  The data consists of thousands of tosses of the magic coin including where the toss took place, who tossed it, who reported the result as well as the time of the day.

```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
toss<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Tosses2022.csv") #web load
# head(moody)
temp<-knitr::kable(
  toss[sample(1:nrow(toss),5), ], caption = 'Snippet of Magic Coin Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```

<br>
```{r,tut=TRUE,height=600}
toss<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Tosses2022.csv")

summary(toss)
```

## What is a good party? Puzzle

Download: <a href="https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Partyb.csv" download="Partyb.csv">Partyb.csv</a>

We store data about thousands of parties, some fun parties, others which were OK or simply boring.  Your goal is to discover secrets of a fun party. Is it music? Dancing? Does who is the host matter? Or who was present at a party? Maybe who was NOT present at the party? We have all this information. 

```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
party<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Partyb.csv") #web load
# head(moody)
temp<-knitr::kable(
  party[sample(1:nrow(party),5), ], caption = 'Snippet of Party Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```

<br>
```{r,tut=TRUE,height=600}
party<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Partyb.csv")

summary(party)
```

## Voter preferences data puzzle

Download: <a href="https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Voting1.csv" download="Voting1.csv">Voting1.csv</a>

```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
voting<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Voting1.csv") #web load
# head(moody)
temp<-knitr::kable(
  voting[sample(1:nrow(voting),5), ], caption = 'Snippet of Voting Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```

<br>
```{r,tut=TRUE,height=600}
vote<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Voting1.csv")

summary(vote)
```

## Good sleep Data Puzzle

Download: <a href="https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/SleepPrediction2.csv" download="SleepPrediction2.csv">SleepPrediction2.csv</a>

```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/SleepPrediction2.csv") #web load
# head(moody)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Sleep Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```

<br>
```{r,tut=TRUE,height=600}
sleep<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/SleepPrediction2.csv")

summary(sleep)
```

## Movies Data Hunt

```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Movies2022F-4.csv") #web load
# head(moody)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Movies Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```
<br>
**Puzzle description:**

contains  imdb scores of 12,800+ movies  along with several attributes including budget, gross  genre, content rating etc.

What are the most promising  alternative hypotheses  about imdb scores to test?  Name your three top candidates along with the evidence which backs them up:  either in the form of R instruction(s)  or plot.

### Secrets Revealed- Patterns in Movies data? 

<button class="btn btn-primary" data-toggle="collapse" data-target="#MV12">Secrets Revealed</button>
<div id="MV12" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/profhw4.pdf&embedded=true" width="100%" height="500px"></embed>
</div>

<!--

### Best Student's Submissions 2022

<button class="btn btn-primary" data-toggle="collapse" data-target="#hw41">Joshua Sze</button>
<button class="btn btn-primary" data-toggle="collapse" data-target="#hw42">Andrew Fasano</button>

<div id="hw41" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/student1hw4.pptx&embedded=true" width="100%" height="500px"></embed>
</div>

<div id="hw42" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/studen2hw4.pptx&embedded=true" width="100%" height="500px"></embed>
</div>
-->

## Professor Canvas data puzzle

You are all familiar with Canvas, right?  This is where you look to see your grades for each assignment and exam.  This is where you see the scores.  In fact, who needs the professor?  Seems like the Professor  is in fact Canvas itself.  However it seems that Professor Canvas went a bit wild and unfair!  

One can still fail the class with the score of 82 (sounds familiar, yes, Professor Moody would do it, but professor Canvas?   How can one get a lower grade  with a higher score?  

Yes, Canvas was instructed by someone and your goal is to discover the grading method. We know who that someone is…it is DataMaker of course. 


## Minimarket Data Hunt 

```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
movies<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/Minimarket.csv") #web load
# head(moody)
temp<-knitr::kable(
  movies[sample(1:nrow(movies),5), ], caption = 'Snippet of Minimarket Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```
<br>
What items sell together? 

A small local minimarket chain (think Wawa at its early days) has a few locations in New Jersey and it sells beer, snacks, sweets. DataMaker provided the data set of several thousand of transactions in the minimarket storing what items were purchased, when they were purchased (weekday or weekend) at which location.  

### What were the secret associations  between  items in the minimarket?

<button class="btn btn-primary" data-toggle="collapse" data-target="#MV12">Secrets Revealed</button>
<div id="MV12" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/ecrets-revealed-prediction-minimarket.pptx&embedded=true" width="100%" height="500px"></embed>
</div>

## Virus Data Puzzle

Whose computer  is in most danger of being infected with a virus?  Let's assume that obvious protections have been secured: antivirus software is installed and an uncrackable password is in place. 

Anything else?  But how about your personal characteristics - are you an optimist or a pessimist?  Do you have lots of facebook friends?  Is your instagram followed?  Massively?  Do you spend all your time online? Are you an outdoorsy type ? Do you have a dog? 

## Good mood Data Puzzle

What are the secrets of DataMaker’s good mood?  Now we fully admit that DataMaker is moody!  Data is not surprisingly very nerdy - amount of free disk space, number of processes, CPU utilization, number of active connections.

## Airbnb data puzzle

## Predicting grades in Professor Moody's class

### How did I cook the Professor Moody Prediction challenge data? 

<button class="btn btn-primary" data-toggle="collapse" data-target="#MV13">Secrets Revealed</button>
<div id="MV13" class="collapse">    
<embed src="https://docs.google.com/viewer?url=https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/pred/Secrets-revealed-prediction-challenge.pptx&embedded=true" width="100%" height="500px"></embed>
</div>

## Addiotional Reference 

<button class="btn btn-primary" data-toggle="collapse" data-target="#p12"> Prediction - Free Style </button> 
<div id="p12" class="collapse">
<embed src="https://docs.google.com/presentation/d/1pA0bzMGr_Tu2CXsgtT9Ks5apHqxWh8l1XXBVvwWr6zc/edit?usp=sharing" width="100%" height="500px"></embed>
</div>
