# {-}
<!-- <script src=https://cdn.datacamp.com/datacamp-light-latest.min.js></script> -->
<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>


```{r ,include=FALSE}
tutorial::go_interactive( greedy = FALSE)

```
This is a textbook based on the [DATA 101](http://data101.cs.rutgers.edu/) course taught by Prof. Tomasz Imielinski at Computer Science Department at Rutgers University, New Brunswick.

Data 101 is an introductory course for beginners from any field of study, interested in the field of Data Science. 

This book's pages are created using [Rmarkdown](https://rmarkdown.rstudio.com/) from Rstudio (similar to jupyter notebook) and the book is compiled using ["bookdown"](https://cran.r-project.org/web/packages/bookdown/index.html) package.

The most important aspect of this interactive book are the interactive code chunks for running code, which are powered by a minimal version of Datacamp's learning interface called "Datacamp Light". For more info visit [Datacamp Light](https://github.com/datacamp/datacamp-light)

- **Note1**: This book is undergoing constant updates following along with the course thought in Spring 2021. Topics under progress are marked with a " * ".

- **Note2**: This book uses Datacamp Light for supporting runnable code chunks. In case the code chunks do not connect to run-time session, please copy the code and run in RStudio. Also, please report if facing this issue to the instructor via [email](mailto:timielinski@gmail.com).

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

