# Lecture: Setting Up R 

- **Important Instructions**
  - Installation of R is required before installing RStudio
    - "R” is a programming language, and,
    - “RStudio” is an Integrated Development Environment (IDE) which provides you a platform to code in R.

- __How to download and install R & RStudio?__

  - _Downloading and installing R._

    - For Windows Users.
      - Click on the link provided below or copy paste it on your favourite browser and go to the website.
          - [https://cran.r-project.org/bin/windows/base/](https://cran.r-project.org/bin/windows/base/)
      - Click on the link at top left where it says “Download R 4.0.3 for windows” or the latest at the time of your installation.
      - Open the downloaded file and follow the instructions as it is.

    - For MAC Users.
      - Click on the link provided below or copy paste it on your favourite browser and go to the website.
          - [https://cloud.r-project.org/bin/macosx/](https://cloud.r-project.org/bin/macosx/)
      - Under “Latest release”, click on “R-4.0.3.pkg” or the latest at the time of your installation.
      - Open the downloaded file and follow the instructions as it is.
      
 
  - _Downloading and installing RStudio._
  
    - For Windows Users.
      - Click on the link below or copy paste it in your favourite browser.
          - [https://rstudio.com/products/rstudio/download/](https://rstudio.com/products/rstudio/download/)
      - Scroll down almost till the end of the web page until you find a section named “All Installers”.
      - Click on the download link beside “Windows 10/8/7” to download the windows version of RStudio.
      - Install RStudio by clicking on the downloaded file and following the instructions as it is.

    - For MAC Users.
      - Click on the link below or copy paste it in your favourite browser.
          - [https://rstudio.com/products/rstudio/download/](https://rstudio.com/products/rstudio/download/)
      - Scroll down almost till the end of the web page until you find a section named “All Installers”.
      - Click on the link beside “macOS 10.13+” to start your download the MAC version of RStudio.
      - Install RStudio by clicking on the downloaded file and following the instructions as it is.

---

## How to upload a data set?
<script src="files/js/dcl.js"></script>
```{r ,include=FALSE}
tutorial::go_interactive()

```

 - To upload the dataset/file present in csv format the read.csv() and read.csv2() functions are frequently used The read.csv() and read.csv2() have different separator symbol: for the former this is a comma, whereas the latter uses a semicolon.

- There are two options while accessing the dataset from your local machine:
  1. To avoid giving long directory paths for accessing the dataset, one should use the command **getwd()** to get the current working directory and store the dataset in the same directory. To access the dataset stored in the same directory one can use the following: **read.csv("moody2020.csv")**
  2. One can also store the dataset at a different location and can access it using the following command: (Suppose the dataset is stored inside the folder Data101_Tutorials on the desktop)
    - For Windows Users.
      - Example: **read.csv("C:/Users/Desktop/Data101_Tutorials/moody2020.csv")**

    - For MAC Users.
      - Example: **read.csv("/Users/Desktop/Data101_Tutorials/moody2020.csv")**
          
**Note: **
The directory path given here is the current working directory hosted on *Github* where the dataset has been stored.
```{r, tut=TRUE}

# Read in the data
df <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

# Print out `df`
head(df)
```

  
