# 🔖 Data Transformation

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive()
```

Very often data is transformed before it is visualized. In this section we review basic transformation techniques. Before we do this, we review a few basic functions which are going to be very useful when transforming data. 

## Basic Functions

### mean() {#mean}

- **mean()** function is used to find the average of values in a numerical vector.

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

#Lets look at the mean of score column.
mean(moody$score)
```


### length()

- **length()** function is used to get the number of elements in any vector

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

#Lets look at the length of the grade column 
length(moody$grade)
```

### max()

- **max()** function is used to get the maximum value in a numerical vector.

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

#lets look at the maximum value of the score in the score column
max(moody$score)
```

### min()

- **min()** function is used to get the minimum value in a numerical vector

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

#Lets look at the minimum value of score in the score column.
min(moody$score)

```

### sd() {#sd}

- **sd()** function is used to find the standard deviation of numerical vector

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

#Lets look at the standard deviation of score column
sd(moody$score)
```

---

Now we are ready to introduce basic data transformation techniques such as slicing and dicing.  Slicing otherwise known as **subsetting** allows to select data frame subsets. These subsets are defined by boolean conditions built from Attribute op value pairs where op is one of the arithmetic operators such as =, !=, < etc.  For example (Score >70)& (Grade ==’A’) refers to subset of a data frame describing students who scored more than 70 points and got an A.  

Dicing refers to eliminating some of the attributes from a data frame  - it is vertical slicing - which results in more “narrow” frame.
Finally we can also expand our data frame with new, so called derived, attributes. This is very useful operation in data analysis since it allows so called **“feature engineering”**.This new user defined features can lead to totally new insights into the data. 

## Subset {#subset}

### Snippet 1- example of subset function {#nrow}

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")
#Subset of rows
moody_never_smartphone<-subset(moody,ON_SMARTPHONE=="never")
nrow(moody)
nrow(moody_never_smartphone)

```

### Snippet 2- example of  subset  function

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")
#Subset of rows
moody1<-subset(moody,ON_SMARTPHONE=="never")
# You can see only student never on smartphone are in the subset.
table(moody1$ON_SMARTPHONE) 

```

### Snippet 3- subset as subframe 

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")
#Alternate way to subset.
moody2<-moody[moody$ON_SMARTPHONE=="never", ]
# You can see a similar table as above.
table(moody2$ON_SMARTPHONE) 

```

### Snippet 4- subsetting columns

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")
colnames(moody)
#subset of columns
moody3<-subset(moody, select = -c(1))
ncol(moody3)
# You can see the number of columns has been reduced by 1, due to sub-setting without column 1
ncol(moody3)

```

### Snippet 5- sub-setting rows and columns

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")
#Subset of Rows and Columns
moody1<-subset(moody, select = c(2:4), ON_SMARTPHONE == "never")
colnames(moody1)
#Notice that only 3 columns are remaining
dim(moody1)

```

### Code Review
#### What would R say?

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")
moody[moody$SCORE>=90,3]
# What will R say?


# A. Get subset of all columns which contains students who scored more than equal to 90
# B. error
# C. get all score values which are more than equal to 90
# D. get subset of only the grades of students with score greater than equal to 90


```

#### What would R say?

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

moody[moody$SCORE>=80.0 & moody$GRADE =='B',] 
# What will R say?

# A. subset of moody data frame who got B grade.
# B. error.
# C. subset of moody data frame with score greater than 80.
# D. subset of moody data frame with score more than 80 and got B grade.


```

---

One of the most important R instructions is **tapply**. It allows parallel execution of an aggregate function for different values of a categorical variable.

## tapply {#tapply}

- **tapply()** computes a measure (mean, median, min, max, etc..) or a function for each factor variable in a vector. It is a very useful function that lets you create a subset of a vector and then apply some functions to each of the subset.
- tapply(numerical, categorical, aggregagte function)

### Snippet 1-  Example of tapply followed by barplot

```{r,height=700}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")


# To apply tapply() on SCORE factored on ON_SMARTPHONE

moody1<-tapply(moody$SCORE,moody$ON_SMARTPHONE,mean)
moody1 # We can see it calculated mean value of the score by students with respect to their use of phone in class.

barplot(moody1,col = "cyan",xlab = "Labels", ylab = "mean_val",main = "tapply() example 1",las = 2, cex.names = 0.75)#plot

```
<!--
### Snippet 2 (*)

```{r,height=700}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

#Lets factor the grades on on_smartphone as well as grade category.
moody2<-tapply(moody$GRADE,list(moody$ON_SMARTPHONE,moody$GRADE),length)
# We can see it calculated count of the grade of student with respect to their in-class smartphone usage  and grade category.
moody2
barplot(moody2,col=c("red","cyan","orange","blue"),main = "tapply() example 2",beside = TRUE,legend=rownames(moody2))
```
-->
### Code Review
#### What would R say?

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

tapply(moody, GRADE, SCORE, min)
# What will R say?

# A. minimum score for each grade
# B. minimum grade for each score
# C. minimum grade only 
# D. Error.


```
<!--
#### What would R say?

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

tapply(moody$GRADE, list(moody$ON_SMARTPHONE,moody$GRADE), length)
# What will R say?

# A. Total no. of students grade for each values of on_smartphone attribute
# B. mean value of on_smartphone attribute for each grade
# C. mean category of on_smartphone only 
# D. error.
```
-->
---

## Derived Attribute

R allows creating new data frame attributes (columns) “on the fly”.  These are new vectors, which are often defined as functions of existing attributes. Hence, the name - derived  attributes. 

Derived attributes will play an important role  in data exploration as well as in building prediction models. Very often, derived attributes allow discovery of important patterns in data. Similarly, derived attributes may be more predictive than original attributes in the imported data sets.

The term feature engineering is often used in machine learning to describe creation of derived attributes.

### Snippet 1  - Making  new categorical attribute. 
 
The  line 4 initializes the new attribute PF (Pass/Fail) to "Pass".  The line 5 replaces "Pass" by “Fail”  for students who received F. This new attribute, PF, will allow exploratory analysis to find “How to pass Professor Moody’s class”. The answer to this question may be different than then answer to “How to get a good grade in Professor Moody’s class”.


```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

# Cut Example using breaks - Cutting data using defined vector. 
moody$PF<-'Pass'
moody[moody$GRADE=='F',]$PF<-'Fail'

# lets see our added column PF
moody

```



### Cut 

- **cut()** function  divides the range of x into intervals. Provides ability to label intervals as well. It plays important role in defining derived attributes from attributes which are numerical.

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

# Cut Example using breaks - Cutting data using defined vector. 
score1 <- cut(moody$SCORE,breaks=c(0,50,100),labels=c("F","P"))
table(score1)

```

### Code Review
#### What would R say?

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

cut(moody$SCORE, breaks=c(0,25,70,100),labels=c("low", "medium", "high"))
#What would R say?

# A. 5 intervals of attribute score
# B. 3 intervals (0,25) (25,70) (75,100)
# C. 3 categorical values "low", "medium" and "high" for different score intervals
# D. 3 separate datasets with similar score values

```

#### What would R say?

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

output<-cut(moody$SCORE, 5)
summary(output)
#What would R say?

# A. 5 intervals of attribute score of unequal count of elements
# B. 5 intervals of attribute score of equal count of elements
# C. 5 categorical values for different score intervals
# D. 5 separate dataset with similar score values

```

#### What would R say?

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

output<-cut(moody$ASKS_QUESTIONS, 2)
summary(output)
#What would R say?

# A. 2 intervals of attribute ask_questions of unequal count of elements in each interval
# B. 2 intervals of attribute ask_questions of equal count of elements in each interval
# C. 2 categorical values for different ask_questions intervals
# D. Error.

```

#### More complex  example of defining derived attributes

The next snippet  illustrates defining a new numerical attribute, $adjustedScore of a student in the Moody data frame.  

Score is adjusted by the value of participation attribute in the following way:   

- If participation is larger than 0.5 - a bonus proportional to participation * 10 is added to the score.  

- If participation is smaller than 0.5, a penalty of 1-participation) * 10 is subtracted from the score. 

In this way, for someone with very small participation, the 10 point penalty will be imposed (10 points subtracted from the score). Conversely,  someone with perfect participation (1.0) will receive a 10 point bonus. 

##### Snippet 1

```{r,height=700}

moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")


moody$conditional <-0
moody[moody$participation<0.50, ]$conditional <- moody[moody$participation<0.50, ]$score -10*(1-moody[moody$participation<0.50, ]$participation)
moody[moody$participation>=0.50, ]$conditional <- moody[moody$participation>=0.50, ]$score +10*moody[moody$participation>=0.50, ]$participation

# print the column names
colnames(moody)

# lets look at the conditional attribute 
head(moody)

#subset the moody dataset rows = 1 to 10 and cols = 1,5
moody[1:10, c(1,5)]

#subset the moody dataset rows = 1 to 10 and cols = 1,5,6
moody[1:10, c(1,5,6)]

# print summary of inidividual columns
summary(moody$score)
summary(moody$conditional)

# Plotting the conditional attribute using boxplot
boxplot(moody$conditional,col = c("red"),main="Complex Example")

# Plotting the score attribute using boxplot
boxplot(moody$score,col = c("blue"),main="Complex Example")

```

## Additional reference

<button class="btn btn-primary" data-toggle="collapse" data-target="#subset12">Data Transformation</button>
<div id="subset12" class="collapse">    
<embed src="https://docs.google.com/presentation/d/1--AeLFVrESqyWO7iNiIBRW7pX-VSYtBco1o8bOpOatE/edit?usp=sharing" width="100%" height="500px"></embed>
</div>