# Subsetting

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive()
```

- **Lecture slides: **     <button class="btn btn-primary" data-toggle="collapse" data-target="#subset12">Subsetting</button>
<div id="subset12" class="collapse">    
<embed src="https://docs.google.com/presentation/d/1L7ml_mwV7ms3eZ2qbznGgNdarhTfrSP6PWq-RxNUgQ/edit?usp=sharing" width="100%" height="500px"></embed>
</div>

## Basic Functions

### mean()

- **mean()** function is used to find the average of the data.

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

#Lets look at the mean of score column.
mean(moody$score)
```


### length()

- **length()** function is used to get or set the length of data.

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

#Lets look at the length of the grade column 
length(moody$grade)
```

### max()

- **max()** function is used to get the maximum valued element in the data.

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

#lets look at the maximum value of the score in the score column
max(moody$score)
```

### min()

- **min()** function is used to get the minimum valued element in the data.

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

#Lets look at the minimum value of score in the score column.
min(moody$score)

```

### sd()

- **sd()** fucntion is used to find the standard deviation of the data.

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

#Lets look at the standard deviation of score column
sd(moody$score)
```

---

## Subset {#subset}

### Snippet 1

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")
#Subset of rows
moody_never_smartphone<-subset(moody,ON_SMARTPHONE=="never")
nrow(moody)
nrow(moody_never_smartphone)

```

### Snippet 2

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")
#Subset of rows
moody1<-subset(moody,ON_SMARTPHONE=="never")
# You can see only student never on smartphone are in the subset.
table(moody1$ON_SMARTPHONE) 

```

### Snippet 3

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")
#Alternate way to subset.
moody2<-moody[moody$ON_SMARTPHONE=="never", ]
# You can see a similar table as above.
table(moody2$ON_SMARTPHONE) 

```

### Snippet 4

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")
#subset of columns
moody3<-subset(moody, select = -c(8))
ncol(moody3)
# You can see the number of columns has been reduced by 1, due to subsetting without column 8
ncol(moody3) 

```

### Snippet 5

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")
#Subset of Rows and Columns
moody4<-subset(moody, select = -c(8), ON_SMARTPHONE == "never")
table(moody4$ON_SMARTPHONE)
dim(moody4)
# You can see only student never on smartphones without column 8 data are present in the subset.
dim(moody4)

```

### Code Review
#### What would R say?

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")
moody[moody$SCORE>=90,3]
# What will R say?


# A. Get subset of all columns which contains students who scored more than equal to 90
# B. error
# C. get all score values which are more than equal to 90
# D. get subset of only the grades of students with score greater than equal to 90


```

#### What would R say?

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

moody[moody$SCORE>=80.0 & moody$GRADE =='B',] 
# What will R say?

# A. subset of moody data frame who got B grade.
# B. error.
# C. subset of moody data frame with score greater than 80.
# D. subset of moody data frame with score more than 80 and got B grade.


```

---

## tapply {#tapply}

- **tapply()** computes a measure (mean, median, min, max, etc..) or a function for each factor variable in a vector. It is a very useful function that lets you create a subset of a vector and then apply some functions to each of the subset.
- tapply(numerical, categorical, aggregagte function)

### Snippet 1

```{r,height=700}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")


# To apply tapply() on SCORE factored on ON_SMARTPHONE

moody1<-tapply(moody$SCORE,moody$ON_SMARTPHONE,mean)
moody1 # We can see it calculated mean value of the score by students with respect to their use of phone in class.

barplot(moody1,col = "cyan",xlab = "Labels", ylab = "mean_val",main = "tapply() example 1",las = 2, cex.names = 0.75)#plot

```

### Snippet 2 (*)

```{r,height=700}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

#Lets factor the grades on on_smartphone as well as grade category.
moody2<-tapply(moody$GRADE,list(moody$ON_SMARTPHONE,moody$GRADE),length)
# We can see it calculated count of the grade of student with respect to their in-class smartphone usage  and grade category.
moody2
barplot(moody2,col=c("red","cyan","orange","blue"),main = "tapply() example 2",beside = TRUE,legend=rownames(moody2))
```

### Code Review
#### What would R say?

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

tapply(moody, GRADE, SCORE, min)
# What will R say?

# A. minimum score for each grade
# B. minimum grade for each score
# C. minimum grade only 
# D. Error.


```

#### What would R say?

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

tapply(moody$GRADE, list(moody$ON_SMARTPHONE,moody$GRADE), length)
# What will R say?

# A. Total no. of students grade for each values of on_smartphone attribute
# B. mean value of on_smartphone attribute for each grade
# C. mean category of on_smartphone only 
# D. error.
```

