# 🔖 Basic R Intructions

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive()
```

In this section we introduce the absolutely basic R instructions, call it R101, which will be sufficient for the entire data 101 class. This is a very small subset of the entire R. The good news is that using this very small subset of R we can accomplish all coding objectives for data 101! 

The set we present  below is a mix of simple arithmetic aggregate functions such as mean() \@ref(mean), max() \@ref(max), sum() , basic data structures such as vectors and data frames and finally, two  core functions defined for data frames: subset() \@ref(subset), tapply() \@ref(tapply) and table() \@ref(table) function defined on vectors. 


## Vector {#vector}

- A vector is simply a list of items that are of the same type.

<!-- To combine the list of items to a vector, use the **c()** function and separate the items by a comma.
- All of the items inside a vector to are of the same type like numerical or categorical. -->

### Snippet 1

Lets look at example of creating a vector:

```{r,tut=TRUE,height=300}

#Lets create 3 vectors with title, author and year.
color <- c('Red','Blue','Yellow','Green')

#Lets look at how the created vectors look.
color
```

### Snippet 2

Create a vector with numerical values in a sequence, use the **:** operator:

```{r,tut=TRUE,height=300}

#Lets create a vectors with numerical sequence.
year <- 2018:2022

#Lets look at how the created vectors look.
year

```

---

<!--
<br />

- To find out how many items a vector has, use the **length()** function:

```{r,tut=TRUE,height=300}

#Lets create a vectors with categorical values
author <- c('Foreman', 'John', 'Said')

# You can access the vector items by referring to its index number inside brackets []. The first item has index 1, the second item has index 2, and so on:
author[1]

#Lets look at the size of a vectors.
length(author)

```

<br />

- To sort items in a vector alphabetically or numerically, use the **sort()** function and to change the value of a specific item, refer to the **index number**:

```{r,tut=TRUE,height=300}

#Lets create a vectors with categorical values
title <- c('Data Smart','Orientalism','False Impressions','Making Software')

#sorting titles
sort(title)

# change the title 'Data Smart' to 'Smart Data'
title[1] <- "Smart Data"

#Lets look at how the updated vectors look.
title 
```

-->

---

## Data Frames

- Data Frames are data displayed in a format as a table.
<!--
- Data Frames can have different types of data inside it. While the first column can be character, the second and third can be numeric or logical. 

- Following are the characteristics of a data frame.
  - The column names should be non-empty.
  - The row names should be unique.
  - The data stored in a data frame can be of numeric, factor or character type.
  - Each column should contain same number of data items.

Use the **data.frame()** function to create a data frame:-->

### Snippet 1

Data frames will serve as containers of imported data - typically data provided in csv format, like the moody data set above. Snippet 4.21 shows how to populate a data frame using read.csv() instruction. Notice that the moody data frame which is the container for the imported  data set  will automatically inherit attribute names (columns) of the underlying data set.

```{r,tut=TRUE,height=300}

# Load the dataset into the moody variable
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")

# Now lets view the dataframe moody with just 5-6 tuples
head(moody)
```

### Snippet 2 

Get the summary of the dataframe:

```{r,tut=TRUE,height=300}

# Load the dataset into the moody variable
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")

# Use the summary() function to summarize the data from a Data Frame:
summary(moody)
```
<!--
- Just like vectors, you can access specific data **(Slicing)** in dataframes using brackets. 
- But now, instead of just using one indexing vector, we use two indexing vectors: one for the rows and one for the columns. -->

### Snippet 3 

We can select subsets of columns and subsets of rows for a data frame using the following the notation data[rows, columns]:

```{r,tut=TRUE,height=500}

# Load the dataset into the moody variable
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")

# Return row 1
moody[1, ]

# Return column 5
moody[, 5]

# Rows 1:5 and column 2
moody[1:5, 2]

# Give me rows 1-3 and columns 2 and 4 of moody
moody[1:3, c(2:4)]
```

---





## Table {#table}

<!--
- Tables (i.e. frequency distribution table) can be created using **table()** along with some of its variations. 
- To use **table()**, simply add in the variables you want to tabulate separated by a comma. 
- You must reference the variable using **dataset$variable**. 
- By default, missing values are excluded from the counts; if you want a count for these missing values you must specify the argument **useNA=“ifany”** or **useNA=“always”**. -->
### Snippet 1

The below examples show how to use this function:

```{r, tut=TRUE,height=400}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv") #web load

#lets make a table for the grades of students and counts of students for each Grade. 
grades <- table(moody$Grade)

#Joint distribution of grade and major
table(moody$Grade, moody$Major)


```



---

## Basic Functions

```{r,echo=FALSE}
# moody<-read.csv("../files/dataset/moody2020b.csv") #static Load
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")
# head(moody)
temp<-knitr::kable(
  moody[sample(1:nrow(moody),5), ], caption = 'Snippet of moody Dataset',
  booktabs = TRUE
)
temp<-knitr::kable(
  moody[sample(1:nrow(moody),5), ], caption = 'Snippet of moody Dataset',
  booktabs = TRUE
)
library(kableExtra)
kableExtra::scroll_box(temp,width = "100%")
```

### mean() {#mean}

- **mean()** function is used to find the average of values in a numerical vector.

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")

#Lets look at the mean of score column.
mean(moody$Score)
```


### length()

- **length()** function is used to get the number of elements in any vector

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")

#Lets look at the length of the grade column 
length(moody$Grade)
```

### max() {#max}

- **max()** function is used to get the maximum value in a numerical vector.

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")

#lets look at the maximum value of the score in the score column
max(moody$Score)
```

### min()

- **min()** function is used to get the minimum value in a numerical vector

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")

#Lets look at the minimum value of score in the score column.
min(moody$Score)

```

### sd() {#sd}

- **sd()** function is used to find the standard deviation of numerical vector

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")

#Lets look at the standard deviation of score column
sd(moody$Score)
```

---




Now we are ready to introduce basic data transformation techniques such as slicing and dicing.  Slicing, otherwise known as **subsetting**, allows the selection of data frame subsets. These subsets are defined by boolean conditions built from Attribute op value pairs where op is one of the arithmetic operators such as =, !=, < etc.  For example (Score >70)& (Grade ==’A’) refers to a subset of a data frame describing students who scored more than 70 points and got an A.  

Dicing refers to eliminating some of the attributes from a data frame  - it is vertical slicing - which results in a more “narrow” frame.
Finally we can also expand our data frame with new, so called derived, attributes. This is a very useful operation in data analysis since it allows so-called **“feature engineering”**. These new user-defined features can lead to totally new insights into the data. 

## Subset {#subset}

The following snippets demonstrate two ways of subsetting a data frame: first  through explicit function subset() and second through the native sub-data frame notation df[ ].

### Snippet 1 {#nrow}

Subsetting data frame via subset() function

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")
#Subset of rows
moody_psychology<-subset(moody, Major== 'Psychology')
nrow(moody)
nrow(moody_psychology)

```

### Snippet 2

Example of  subset  function

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")

#Alternate way to subset.
moody[moody$Major=="Psychology", ]
moody[moody$Major!="Psychology", ]
moody[moody$Score >80, ]
moody[moody$Score >80 & moody$Grade == 'B', ]

```

### Snippet 3

Subsetting columns

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")
colnames(moody)
#subset of columns
moody3<-subset(moody, select = -c(1))
ncol(moody3)
# You can see the number of columns has been reduced by 1, due to sub-setting without column 1
ncol(moody3)

```

### Snippet 4

Sub-setting rows and columns

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")
#Subset of Rows and Columns
moody1<-subset(moody, select = c(2:4), Major=="Psychology")
colnames(moody1)
#Notice that only 3 columns are remaining
dim(moody1)

```

---



One of the most important R instructions is **tapply**. It allows parallel execution of an aggregate function for different values of a categorical variable.

## tapply {#tapply}

tapply()  has four arguments:  the data frame (df), numerical attribute of df, categorical attribute of df and aggregate function (mean, max, min etc).  Syntax of df is as follows:
 
```tapply(df$numerical attribute, df$categorical attribute, aggregate function)```

- tapply() first slices data frame  df by different values of a categorical attribute and then computes an aggregate  (mean, median, min, max, etc..) of a numerical attribute  to each slice. 


### Snippet 1

Example of tapply followed by plot

```{r,height=700}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")

# To see distributions of mean scores for different grades only for students who were “always” on smartphones in class. 
tapply(moody[moody$Seniority == "Junior",]$Score, moody[moody$Seniority== "Junior",]$Grade, mean)

```

### Snippet 2

Combining table and subset

```{r,height=300}

moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2022.csv")

#Distribution of grades for juniors
table(moody[moody$Seniority == 'Junior',]$Grade)
#Distribution of grades for seniors who major in Economics
table(moody[moody$Seniority == 'Senior' & moody$Major == 'Economics', ,]$Grade)
```



<!--
### Snippet 2 (*)

```{r,height=700}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

#Lets factor the grades on on_smartphone as well as grade category.
moody2<-tapply(moody$GRADE,list(moody$ON_SMARTPHONE,moody$GRADE),length)
# We can see it calculated count of the grade of student with respect to their in-class smartphone usage  and grade category.
moody2
barplot(moody2,col=c("red","cyan","orange","blue"),main = "tapply() example 2",beside = TRUE,legend=rownames(moody2))
```



### Cut 

- **cut()** function  divides the range of x into intervals. Provides ability to label intervals as well. It plays important role in defining derived attributes from attributes which are numerical.

```{r}
moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/MOODY-2019.csv")

# Cut Example using breaks - Cutting data using defined vector. 
score1 <- cut(moody$SCORE,breaks=c(0,50,100),labels=c("F","P"))
table(score1)

```


#### More complex  example of defining derived attributes

The next snippet  illustrates defining a new numerical attribute, $adjustedScore of a student in the Moody data frame.  

Score is adjusted by the value of participation attribute in the following way:   

- If participation is larger than 0.5 - a bonus proportional to participation * 10 is added to the score.  

- If participation is smaller than 0.5, a penalty of 1-participation) * 10 is subtracted from the score. 

In this way, for someone with very small participation, the 10 point penalty will be imposed (10 points subtracted from the score). Conversely,  someone with perfect participation (1.0) will receive a 10 point bonus. 

##### Snippet 1

```{r,height=700}

moody<-read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")


moody$conditional <-0
moody[moody$participation<0.50, ]$conditional <- moody[moody$participation<0.50, ]$score -10*(1-moody[moody$participation<0.50, ]$participation)
moody[moody$participation>=0.50, ]$conditional <- moody[moody$participation>=0.50, ]$score +10*moody[moody$participation>=0.50, ]$participation

# print the column names
colnames(moody)

# lets look at the conditional attribute 
head(moody)

#subset the moody dataset rows = 1 to 10 and cols = 1,5
moody[1:10, c(1,5)]

#subset the moody dataset rows = 1 to 10 and cols = 1,5,6
moody[1:10, c(1,5,6)]

# print summary of inidividual columns
summary(moody$score)
summary(moody$conditional)

# Plotting the conditional attribute using boxplot
boxplot(moody$conditional,col = c("red"),main="Complex Example")

# Plotting the score attribute using boxplot
boxplot(moody$score,col = c("blue"),main="Complex Example")

```
-->

## Additional references

<button class="btn btn-primary" data-toggle="collapse" data-target="#subset12">Data Transformation</button>
<div id="subset12" class="collapse">    
<embed src="https://docs.google.com/presentation/d/1--AeLFVrESqyWO7iNiIBRW7pX-VSYtBco1o8bOpOatE/edit?usp=sharing" width="100%" height="500px"></embed>
</div>