# How to upload a data set?
<script src="files/js/dcl.js"></script>
```{r ,include=FALSE}
tutorial::go_interactive()

```

 - To upload the dataset/file present in csv format the read.csv() and read.csv2() functions are frequently used The read.csv() and read.csv2() have different separator symbol: for the former this is a comma, whereas the latter uses a semicolon.

- There are two options while accessing the dataset from your local machine:
  1. To avoid giving long directory paths for accessing the dataset, one should use the command **getwd()** to get the current working directory and store the dataset in the same directory. To access the dataset stored in the same directory one can use the following: **read.csv("moody2020.csv")**
  2. One can also store the dataset at a different location and can access it using the following command: (Suppose the dataset is stored inside the folder Data101_Tutorials on the desktop)
    - For Windows Users.
      - Example: **read.csv("C:/Users/Desktop/Data101_Tutorials/moody2020.csv")**

    - For MAC Users.
      - Example: **read.csv("/Users/Desktop/Data101_Tutorials/moody2020.csv")**
          
**Note: **
The directory path given here is the current working directory hosted on *Github* where the dataset has been stored.
```{r, tut=TRUE}

# Read in the data
df <- read.csv("https://raw.githubusercontent.com/dev7796/data101_tutorial/main/files/dataset/moody2020b.csv")

# Print out `df`
head(df)
```

  

