# Vector

<script src="files/js/dcl.js"></script>
<link rel="stylesheet" href="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/css/bootstrap.min.css">
<script src="https://maxcdn.bootstrapcdn.com/bootstrap/3.4.1/js/bootstrap.min.js"></script>

```{r ,include=FALSE}
tutorial::go_interactive( greedy = FALSE)

```
- A vector is simply a list of items that are of the same type.
- To combine the list of items to a vector, use the **c()** function and separate the items by a comma.
- All of the items inside a vector to are of the same type like numerical or categorical. 


Lets look at example of creating a vector:
```{r,tut=TRUE,height=300}

#Lets create 3 vectors with title, author and year.
title <- c('Data Smart','Orientalism','False Impressions','Making Software')
author <- c('Foreman, John','Said, Edward','Archer, Jeffery','Oram, Andy')
year <- c(2010,2011,2012,1998)

#Lets look at how the created vectors look.
title
author
year

```

<br />
Create a vector with numerical values in a sequence, use the **:** operator:

```{r,tut=TRUE,height=300}

#Lets create a vectors with numerical sequence.
year <- 1:20

# Vector with numerical decimals in a sequence
weight <- 100.5 : 210.6

# To make bigger or smaller steps in a sequence, use the seq() function
height <- seq(from = 0, to = 100, by = 20)

#Lets look at how the created vectors look.
year
size
height

```

<br />
To find out how many items a vector has, use the **length()** function:

```{r,tut=TRUE,height=300}

#Lets create a vectors with categorical values
author <- c('Foreman', 'John', 'Said')

# You can access the vector items by referring to its index number inside brackets []. The first item has index 1, the second item has index 2, and so on:
author[1]

#Lets look at the size of a vectors.
length(author)

```

<br />
To sort items in a vector alphabetically or numerically, use the **sort()** function and to change the value of a specific item, refer to the **index number**:

```{r,tut=TRUE,height=300}

#Lets create a vectors with categorical values
title <- c('Data Smart','Orientalism','False Impressions','Making Software')

#sorting titles
sort(title)

# change the title 'Data Smart' to 'Smart Data'
title[1] <- "Smart Data"

#Lets look at how the updated vectors look.
title 
```





